Certifying, incremental type checking

### 2012-06-12 doc/papers/LFMTP2012/paper.tex
paper

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item a computational language to write untrusted
  derivation-generating type checkers, based on a notion of function
  inverse %TODO le nom?
  to refer to already-computed values;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated certificates
  incrementally against a \emph{repository} of already constructed and
  checked derivations.
\end{itemize}

@@

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
% TODO: vérif ce paragraphe après avoir parlé de fcts inv
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would arrive to the unhandled case $\finfer{\var
  x}$. This requires evaluating the type-checker with a carefully
designed strategy (see \ref{sec:comput}).

@@

To write deltas by sharing subterms, we need to be able to address any
subterm. Since a subterm can be open, we must instantiate its free
variables. For this, we define the \emph{slicing} of a well-typed \LF\
term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
Y$, \ldots\ as in \cite{nanevski2008contextual}. Metavariables stand
for well-typed terms defined in a repository, and are instantiated
wrt. their free variables. A repository also contains a distinguished
metavariable corresponding to the ``tip'' of the term stored in
it. For instance, the repository
$$
\small
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
is one of the possible way of slicing the term
$$
\fbox{$
  \app {\cst{lam}} \lam x
  \fbox{$
    \app {\cst{lam}} \lam y
    \fbox{$
      \app {\cst{app}} \app {\var y}
      (\fbox{$
        \app{\cst{app}}\app{\var x}{\cst o}
      $})
    $}
  $}
$}
$$
where all boxed subterms are referable slices.

\subsection{The computational calculus}
\label{sec:comput}

Secondly, we need a mean to transform our \LF\ objects as in
\ref{sec:lang-incr}. For this we introduce \emph{interpreted
  functions} $\interp f$, $\interp g$, \ldots in the signatures,
besides constants. Each $\interp f$ comes with its type and its code:

% inverse

@@

The value calculus defines the \emph{values} we want to
manipulate. These contain constants, and --- contrarily to \eg\
\sysname{ML} values --- binders and variables enabling $\lambda$-tree
syntax. Thus, they embed a superficial notion of computation:
hereditary substitution. This calculus is an extension of \emph{spine
  canonical \LF} \cite{pfenning2007term} enriched with metavariables
which are references to open terms as in
\cite{nanevski2008contextual}. We present its typing algorithm
relatively to a signature. % It features an explicit caching mechanism
% using metavariables.

\paragraph{Syntax}

In this variant, only $\eta$-long and $\beta$-normal values have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
\emph{spine} of objects) as opposed to the more standard binary
application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}.

\begin{align*}
  K &\gequal
  \prd x A K \gor
  \type &
  \text{Kind}\\
  A &\gequal
  \prd x A A \gor
  P &
  \text{Type family} \\
  P &\gequal
  \app {\cst a} S &
  \text{Atomic type} \\
  M &\gequal
  \lam x M \gor
  F &
  \text{Canonical object} \\
  F &\gequal \app H S
  \gor
  \smeta X \sigma &
  \text{Atomic object} \\
  H &\gequal
  \var x \gor
  \cst c &
  \text{Head}\\
  S &\gequal
  \spinenil \gor
  \spinecons M S &
  \text{Spine}\\
  \sigma &\gequal
  \msubstnil \gor
  \msubstcons \sigma x M &
  \text{Parallel substitution} \\
  \Gamma &\gequal
  \envnil \gor
  \envcons \Gamma x A &
  \text{Environment}
\end{align*}

As is customary, we adopt the Barendregt convention for naming bound
variables, write $A\to A'$ for $\prd x A A'$ when $\var x\notin \FV{A'}$,
$\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list, map
or set notations for $\sigma$ and $\Gamma$. The identity substitution
$\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var x}\ |\ \var x
\in\dom\Gamma\}$.

Metavariables $\smeta X \sigma$ stand for open objects with free
variables in $\dom{\sigma}$; To each use of them is attached a
substitution $\sigma$ defining these free variables. They are defined
in a repository (see below). We write $\FMV{M}$ for the set of
metavariables in $M$ and speak of a \emph{metaclosed} object $M$ if
$\FMV{M} = \emptyset$.

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), an idea dating back to \cite{pierce2000local}:
canonical objects are checked against a given type, whereas we can
synthesize the type of atomic objects; the coercion from the first
grammatical category to the second is where we verify equality between
the synthesized and the checked types.

\paragraph{Substitution}

We overload \emph{substitution} as a partial operation on objects,
given a parallel substitution $\sigma$. Since the syntax of object is
not stable by (textual) substitution, it is defined as
\emph{hereditary}\footnote{ Note that this notion of substitution,
  present in \cite{pfenning2007term}, differs from the more standard
  definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
  normal forms: once a \emph{cut} starts, there must be exactly as
  many formal and actual arguments. A non-empty spine cut against
  another non-empty spine $\gcut {\app H S} {S'}$ blocks the
  substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
  implicit $\eta$-expansion); conversely, exhausting actual but not
  formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
  substitution too. For example, the reduction $\gcut {\lam x \app
    {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
  because the second $\var x$ in the left part is substituted by a
  function yet is not itself a function. }: it is mutually recursive
with a \emph{cut} function $\gcut M S$ which can in turn trigger a
chain of substitutions, leading to a canonical objects or an error.
\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $\var x\notin \dom{\sigma}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(\var x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons M S}\sigma &=
  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M {\msubst x N}} S
  \\
  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}

We extend this operation trivially on families $\gsubst A \sigma$ and
kinds $\gsubst K \sigma$. We also allow to abusively write
non-canonical application $\app M S$ as a shorthand for $\gcut M S$
if it is defined.

@@

A \emph{signature} declares object and type constants to be used in
objects. To each object constant we assign an annotation $L$ setting
the slicing behaviour of the constant during type checking: either
sliceable $\annsliceable$ (each application of this constant will be
cached and given a name), non-sliceable $\annnonsliceable$ (we don't
create a name for applications of this constant), or defined
$\anndefined{T}$ (see \ref{sec:computational-calculus}).
\begin{align*}
  L &\gequal
  \annsliceable \gor
  \annnonsliceable \gor
  \anndefined{T}
  \\
  \Sigma &\gequal
  \gnil \gor
  \gcons \Sigma {\cst c :^L A} \gor
  \gcons \Sigma {\cst a : K}
  &\text{Signature}
\end{align*}
We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
suggesting the implementation of signatures as maps.

@@

A \emph{repository} $\mr$ is a finite map from metavariable to
judgments of atomic object well-typing, along with a distinguished
metavariable, corresponding to the \emph{tip} of the term contained in
it:
$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P)), \smeta X
\sigma $$
The algorithm will maintain the invariant that a repository
$\mr$ is \emph{acyclic}, that is for all $\meta X$ in $\mr$, the
fixpoint of $\FMV{\meta X}$ does not contain $\meta X$. If $\mr =
(\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$, abusively
$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
if $\Delta(\meta X) = (\Gamma\vdash M : A)$.

We refer to cached subterms by metavariables. We force instantiation
of all variables in their environment by attaching a substitution to
them ($\smeta X \sigma$).

@@

\paragraph{Checkout}

We define the partial \emph{checkout} operation $\checkout{\mr}$
(resp. $\checkoutr{M}$, $\checkoutr{S}$) from a repository
(resp. object, spine) to a metaclosed object as unfolding the
definitions of all metavariables recursively:
\begin{align*}
  \checkout{\mr} &= \checkoutr{\hat{\mr}} \\
  \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
  \checkoutr{\lam x M} &= \lam x \checkoutr{M} \\
  \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
  \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} \\
  \checkoutr{\spinenil} &= \spinenil
\end{align*}

A repository is said to be \emph{well-formed} if $\checkout {\mr}$
is defined.

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects. We have the ability to construct
objects out of other, and deconstruct them by case analysis:
\begin{align*}
  T &\gequal
  \lam x T \gor
  U & \text{Term} \\
  U &\gequal
  F \gor
  \matchin U \Gamma C & \text{Atomic term}\\
  C &\gequal \casebnobar P U \gor
  C\ \caseb P U & \text{Branches}\\
  P &\gequal
  \app H {\var x} \ldots\ {\var x} &\text{Pattern}
\end{align*}

This language is un(i)typed, in that all \LF\ objects are viewed as
having the same implicit type $\cst{obj}$: there is no static
guarantee that terms produced by a constant $\cst c : A_1\to A_2$ will
have type $A_2$. It is also non-terminating: we could declare a
recursive constant $\cst c : \cst a\to \cst a = \lam x \app {\cst c}
{\var x}$. We expose here only the minimal building blocks of this
language, but really any language with algebraic datatypes would do
since we will be manipulating directly the first-order representation
of \LF\ objects. In fact, whatever the language is, you can think of
defined constants as mere black boxes taking and producing terms, and
being able to call other black boxes.

Nevertheless, in order to maintain the well-typing of the generated
term, we want to maintain the invariant that each time a piece of term
goes in or out of a black box, it is checked against its specification
(\ie\ its type). We thus maintain \emph{dynamic typing} of
computations: arguments and results are type-checked at run-time.

\paragraph{Substitution}

>>>>

In this paper, we present a framework to incrementally generate typing
certificates represented in the \LF\ logical framework, and implemented
as an \sysname{ML} library. We specifically propose:
\begin{itemize}
\item a simple language for representing programs and proof
  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
\item a computational language to write untrusted
  derivation-generating type checkers, based on a notion of function
  inverse %TODO le nom?
  to refer to already-computed values;
\item a generic algorithm to evaluate these type-checkers on program
  deltas and verify on-the-fly the generated incremental certificates
  against a \emph{repository} of already constructed and checked
  derivations.
\end{itemize}

@@

Introducing the ``inverse'' of the derivation inference function has
two purposes: as we saw in \ref{sec:use-incremental}, it allows to
refer to already-constructed derivations when writing deltas. It also
permits to abstract from the concrete environment when writing the
type-checker itself, that one must thread throughout the process in a
traditional, first-order type-checker. It is a generalization of the
so-called \emph{context-free} typing presented in
\cite[chap. 4]{boespflug2011conception} where a special variable
construct $\{x:A\}$ denoting free variables is added to the syntax of
terms, and the rules
$$
\infer{
  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
}{
  \vdash \tlam x A M : A\to B
}
\qquad\text{and}\qquad
\infer{ }{
  \vdash \{\var x : A\} : A
}
$$
% TODO: vérif ce paragraphe après avoir parlé de fcts inv
replace the usual ones. The difficulty with this technique,
particularly in a dependent setting, is to ensure that term $\{\var
x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
can strip off the annotations, while not forgetting them too soon:
otherwise we would hit the unhandled case $\finfer{\var x}$. This
requires evaluating the program deltas with a carefully designed
strategy (see \ref{sec:comput}).

@@

To write deltas by sharing subterms, we need to be able to address any
subterm. Since a subterm can be open, we must instantiate its free
variables. For this, we define the \emph{slicing} of a well-typed \LF\
term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
Y$, \ldots\ as in \cite{nanevski2008contextual}. Metavariables stand
for well-typed terms defined in a \emph{repository}, and are
instantiated wrt. their free variables. A repository also contains a
distinguished metavariable corresponding to the ``tip'' of the term
stored in it.

We thus introduce Sliced \LF\ (\SLF, see
Fig. \ref{fig:syntax-LF}). Specifically, it is an extension of
\emph{spine canonical \LF} \cite{pfenning2007term} enriched with
metavariables which are references to open terms as in
\cite{nanevski2008contextual}.

\begin{figure}\centering
  \begin{align*}
    K &\gequal \prd x A K \gor \type &
    \text{Kind}\\
    A &\gequal \prd x A A \gor P &
    \text{Type family} \\
    P &\gequal \app {\cst a} S &
    \text{Atomic type} \\
    M &\gequal \lam x M \gor F &
    \text{Canonical object} \\
    F &\gequal \app H S \gor \smeta X \sigma &
    \text{Atomic object} \\
    H &\gequal \var x \gor \cst c &
    \text{Head}\\
    S &\gequal \spinenil \gor \spinecons M S &
    \text{Spine}\\
    \sigma &\gequal \msubstnil \gor \msubstcons \sigma x M &
    \text{Parallel substitution} \\
    \Gamma &\gequal \envnil \gor \envcons \Gamma x A &
    \text{Environment}
  \end{align*}
  \caption{Syntax of \SLF}
  \label{fig:syntax-LF}
\end{figure}

In this variant, only $\eta$-long and $\beta$-normal values have an
existence: canonicity for $\beta$-reduction is enforced in the syntax
and canonicity for $\eta$-expansion will be enforced by typing,
following \cite{hl07mechanizing}. Besides, application of a
\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
\emph{spine} of objects) as opposed to the more standard binary
application: this eases the definition substitution and
$\eta$-expansion, and corresponds to a focused sequent calculus known
as \sysname{LJT} \cite{herbelin1995λ}. The definition of hereditary
substitution $\gsubst M \sigma$ is skipped and can be found in
\cite{pfenning2007term} (it is extended to metavariables as in
\cite{nanevski2008contextual}).

We distinguish syntactically between \emph{checkable} $M$ and
\emph{inferrable} $F$ objects (resp. \emph{canonical} and
\emph{atomic}), as in \eg\ \cite{pierce2000local}:
canonical objects are checked against a given type, whereas we can
synthesize the type of atomic objects; the coercion from the first
grammatical category to the second is where we verify equality between
the synthesized and the checked types.

As is customary, we adopt the Barendregt convention for naming bound
variables, write $A\to A'$ for $\prd x A A'$ when $\var x\notin
\FV{A'}$, $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently
list, map or set notations for $\sigma$ and $\Gamma$. The identity
substitution $\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var
  x}\ |\ \var x \in\dom\Gamma\}$. Metavariables $\smeta X \sigma$
stand for open objects with free variables in $\dom{\sigma}$; To each
use of them is attached a substitution $\sigma$ defining these free
variables. We write $\FMV{M}$ for the set of metavariables in $M$ and
speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.

% \paragraph{Substitution}

% We overload \emph{substitution} as a partial operation on objects,
% given a parallel substitution $\sigma$. Since the syntax of object is
% not stable by (textual) substitution, it is defined as
% \emph{hereditary}\footnote{ Note that this notion of substitution,
%   present in \cite{pfenning2007term}, differs from the more standard
%   definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
%   normal forms: once a \emph{cut} starts, there must be exactly as
%   many formal and actual arguments. A non-empty spine cut against
%   another non-empty spine $\gcut {\app H S} {S'}$ blocks the
%   substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
%   implicit $\eta$-expansion); conversely, exhausting actual but not
%   formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
%   substitution too. For example, the reduction $\gcut {\lam x \app
%     {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
%   because the second $\var x$ in the left part is substituted by a
%   function yet is not itself a function. }: it is mutually recursive
% with a \emph{cut} function $\gcut M S$ which can in turn trigger a
% chain of substitutions, leading to a canonical objects or an error.
% \begin{align*}
%   \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
%   \text{if $\var x\notin \dom{\sigma}$}
%   \\
%   \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
%     \sigma)}
%   \\
%   \gsubst {(\app {\var x} S)} \sigma &=
%   \app {\var x} {\gsubst S \sigma} &
%   \text{if $\var x\notin \dom{\sigma}$}
%   \\
%   \gsubst {(\app {\var x} S)} \sigma &=
%   \gcut M {\gsubst S \sigma} &
%   \text{if $\sigma(\var x) = M$}
%   \\
%   \gsubst {\smeta X {\sigma}} {\sigma'} &=
%   \smeta X {\gcomp\sigma{\sigma'}}
%   \\
%   \gsubst\spinenil\sigma &=
%   \spinenil \\
%   \gsubst{\spinecons M S}\sigma &=
%   \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
%   \\
%   \gcut {\lam x M} {\spinecons N S} &=
%   \gcut {\gsubst M {\msubst x N}} S
%   \\
%   % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
%   % \gcut {\smeta X\sigma} S &=
%   % \smeta X{\gcomp\sigma\sigma'}
%   % \\
%   \gcut {\app H S} \spinenil &=
%   \app H S
%   \\
%   \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
%   \\
%   \gcomp \sigma \msubstnil &=
%   \sigma
%   \\
%   \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
%   \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
%   \\
%   \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
%   \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
%   \\
%   \gsubst {\msubstnil}{\sigma} &= \msubstnil
% \end{align*}

% We extend this operation trivially on families $\gsubst A \sigma$ and
% kinds $\gsubst K \sigma$. We also allow to abusively write
% non-canonical application $\app M S$ as a shorthand for $\gcut M S$
% if it is defined.

\begin{definition}
  A \emph{repository} $\mr$ is a finite map from metavariable to
  judgments of atomic object well-typing, along with a distinguished
  metavariable, corresponding to the \emph{tip} of the term contained
  in it:
$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P)), \smeta X
\sigma $$
\end{definition}

The algorithm will maintain the invariant that a repository $\mr$ is
\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint of
$\FMV{\meta X}$ does not contain $\meta X$. If $\mr = (\Delta,
X[\sigma])$, we write $\hat {\mr}$ for the tip $X[\sigma]$, abusively
$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
if $\Delta(\meta X) = (\Gamma\vdash M : A)$.

\begin{definition}
  We define the partial \emph{checkout} operation $\checkout{\mr}$
  (resp. $\checkoutr{M}$, $\checkoutr{S}$) from a repository
  (resp. object, spine) to a metaclosed object as unfolding the
  definitions of all metavariables recursively:
  \begin{align*}
    \checkout{\mr} &= \checkoutr{\hat{\mr}} \\
    \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
    \checkoutr{\lam x M} &= \lam x \checkoutr{M} \\
    \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
    \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} \\
    \checkoutr{\spinenil} &= \spinenil
  \end{align*}
\end{definition}

\begin{example}
  The repository
$$
\small
\left(\begin{array}{llll}
    \meta X &\mapsto
    \envnil&\vdash
    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
    \cst{tm}
    \\
    \meta Y &\mapsto
    \env u {\cst{tm}} &\vdash
    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
      {\var u}} v {\var y} } &:
    \cst{tm}
    \\
    \meta Z &\mapsto
    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
    \app{\app{\cst{app}}{\var v}}{\var u} &:
    \cst{tm}
    \\
    \meta T &\mapsto
    \env x {\cst{tm}} &\vdash
    \app{\cst{app}}\app{\var x}{\cst{o}} &:
    \cst{tm}

\end{array}\right), \smeta X {}
$$
checks out as the metaclosed atomic object
$$
\fbox{$
  \app {\cst{lam}} \lam x
  \fbox{$
    \app {\cst{lam}} \lam y
    \fbox{$
      \app {\cst{app}} \app {\var y}
      (\fbox{$
        \app{\cst{app}}\app{\var x}{\cst o}
      $})
    $}
  $}
$}
$$
where all referable slices have been boxed.
\end{example}

\subsection{The computation language}
\label{sec:comput}

Secondly, we introduce the \CL\ language to transform our \SLF\
objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
other, and deconstruct them by case analysis on an applicative
term. These objects are \SLF\ objects, except that they can contain
special function symbols $\fct f$ and $\fcti f n$. Hence, we name them
with underlined metavariables $\cl F$ instead of $F$.

\begin{figure}
  \centering
  \begin{align*}
    T &\gequal \lam x T \gor
    U & \text{Term} \\
    U &\gequal {\cl F} \gor
    \matchin U \Gamma C & \text{Atomic term}\\
    C &\gequal \casebnobar P U \gor
    C\ \caseb P U & \text{Branches}\\
    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern} \\
    \cl M &\gequal \lam x {\cl M} \gor \cl F &
    \text{Canonical object} \\
    \cl F &\gequal \app {\cl H} {\cl S} \gor \smeta X \sigma &
    \text{Atomic object} \\
    \cl H &\gequal \var x \gor \cst c \gor \fct f \gor \fcti f n&
    \text{Head}\\
    \cl S &\gequal \spinenil \gor \spinecons {\cl M} {\cl S} &
    \text{Spine}\\
    \cl\sigma &\gequal \msubstnil \gor \msubstcons {\cl\sigma} x {\cl M} &
    \text{Parallel substitution} \\
  \end{align*}

  \caption{Syntax of \CL}
\label{fig:syntax-CL}
\end{figure}

This language is dynamically typed: there is no static guarantee that
object $\app {\fct f} x$ produced by a function $\fct f : A_1\to A_2 =
T$ will have type $A_2$, nor that pattern matching is exhaustive. It
is also non-terminating: we can define a recursive function $\fct f :
\cst a\to \cst a = \lam x \app {\fct f} {\var x}$. Nevertheless, we
want to maintain the invariant that each time an object goes in or out
of a function $\fct f : A_1\to A_2$, it is checked against the type of
this function (resp. $A_1$ or $A_2$). You can think of interpreted
functions as mere black boxes taking and producing objects, and being
able to feed objects to other black boxes; yet the wires linking them
are labeled with types and objects circulating in them are checked
\wrt\ that type.

Besides constants, the signature defines \emph{interpreted functions}
$\fct f$, $\fct g$. Each function comes with its type and its
code.

\begin{definition}
  A \emph{signature} declares object and type constants and functions
  to be used in objects.
  \begin{align*}
    \Sigma &\gequal
    \gnil
    \gor
    \gcons \Sigma {\cst a : K}
    \gor
    \gcons \Sigma {\cst c : A}
    \gor
    \gcons \Sigma {\fct f : A = T}
    &\text{Signature}
  \end{align*}
\end{definition}

Again, we use indifferently set, map or list notations for signatures.

\begin{definition}
  To each function $\fct f : A = T$ in $\Sigma$ we associate a family
  of \emph{inverses} $\fcti f n : (A)^n = \pi_n(A)$ for $n\in\mathbb
  N$, where type $(A)^n$ and term $\pi_n(A)$ are defined recursively:
  \begin{align*}
    (\prd x {A_1} {A_2})^{n+1} &= \prd x {A_1} (A_2)^n &
    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)^{A_1} \\
    (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
    (\app {\cst a} S)^A &= \prd x {\app{\cst a} S} A \\
    \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_n) \\
    \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
    \pi^y(\app{\cst a} S) &= \var y
  \end{align*}
\end{definition}

\begin{example}
  Function $\fct{infer} : (\prd M {\cst{tm}} \app{\cst{sig}} M) =
  T$ (called $\finfer{}$ above, with $\cst{sig}$ the \LF\ encoding of
  the previous $\Sigma$-type) has one inverse $\fcti{infer}{0} :
  (\prd M {\cst{tm}} {\app{\cst{sig}} M} \to \cst{tm}) = \lam x \lam y
  \var x$ (called $\fget{}$ above).
\end{example}

\begin{example}
  Function $\fct{equal} : \prd M {\cst{tm}} \prd N {\cst{tm}}
  \app{\cst{eq}}\app M N = T'$ has two inverses: $\fcti{equal}{0}$ and
  $\fcti{equal}{1}$ of the same type $\prd M {\cst{tm}} \prd N
  {\cst{tm}} \app{\cst{eq}}\app M N \to \cst{tm}$ with code
  respectively $\lam m \lam n \lam h \var m$ and $\lam m \lam n \lam h
  \var n$.
\end{example}

@@

% The value calculus defines the \emph{values} we want to
% manipulate. % These contain constants, and --- contrarily to \eg\
% \sysname{ML} values --- binders and variables enabling $\lambda$-tree
% syntax. Thus, they embed a superficial notion of computation:
% hereditary substitution.

@@

@@

% We refer to cached subterms by metavariables. We force instantiation
% of all variables in their environment by attaching a substitution to
% them ($\smeta X \sigma$).

@@

% A repository is said to be \emph{well-formed} if $\checkout {\mr}$
% is defined.

@@

We now turn to the definition of our computation calculus. As we saw,
code is attached to special, \emph{defined} constants in the signature
that manipulates \LF\ objects.
\paragraph{Substitution}
